---
title: "overview_analysis"
author: "Connor Flynn"
date: "12/8/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(readr)
library(janitor)
library(tidyr)
library(tidyverse)
library(plotly)
library(RColorBrewer)
library(dplyr)
library(forcats)
library(hrbrthemes)
library(viridis)

options(scipen = 99)

```

The purpose of this R Markdown is to perform a preliminary, overview analysis of the dataset. 


Import Data


```{r}
nmdata <- read_csv(here("data/Literature_transcriptomics_review_and_data_of_Nanoparticle_Induced_Cellular_Outcomes.csv"))
```

Clean Column Names

```{r}
nmdata <- nmdata %>% 
  clean_names()
```


```{r}

names(nmdata) <- gsub(x = names(nmdata), 
                                        pattern = "unfolded_protein_responses_upr_and_endoplasmatic_reticulum_er_stress", 
                                        replacement = "UPR_ER_Stress")

names(nmdata) <- gsub(x = names(nmdata), 
                                        pattern = "angiogenenesis_responses", 
                                        replacement = "Angiogenesis")

names(nmdata) <- gsub(x = names(nmdata), 
                                        pattern = "metal_ion_responses", 
                                        replacement = "Metal_Ion")

names(nmdata) <- gsub(x = names(nmdata), 
                                        pattern = "inflammation_and_immune_responses", 
                                        replacement = "Inflammation_Immune")

names(nmdata) <- gsub(x = names(nmdata), 
                                        pattern = "cell_adhesion_and_extracellular_matrix_responses", 
                                        replacement = "Cell_Adhesion_ECM")

names(nmdata) <- gsub(x = names(nmdata), 
                                        pattern = "dna_damage_and_repair_responses", 
                                        replacement = "DNA_Damage_Repair")

names(nmdata) <- gsub(x = names(nmdata), 
                                        pattern = "cell_death_and_apoptosis_responses", 
                                        replacement = "Cell_Death_Apoptosis")

names(nmdata) <- gsub(x = names(nmdata), 
                                        pattern = "cell_cycle_and_proliferation_responses", 
                                        replacement = "Cell_Cycle_and_Proliferation")

names(nmdata) <- gsub(x = names(nmdata), 
                                        pattern = "cytoskeleton_organization_responses", 
                                        replacement = "Cytoskeleton_Organization")


  
```


```{r}
nmdata[nmdata == "Ag_NPs"] <- "Ag"
nmdata[nmdata == "Au_NPs"] <- "Au"

nmdata[nmdata == "CuO_NPs"] <- "CuO"
nmdata[nmdata == "Fe2O3_NPs"] <- "Fe203"
nmdata[nmdata == "Polymeric_NPs"] <- "Polymeric"
nmdata[nmdata == "Polystyrene_NPs"] <- "Polystyrene"
nmdata[nmdata == "TiO2_NPs"] <- "TiO2"
nmdata[nmdata == "ZnO_NPs"] <- "ZnO"
```





Select Nanoparticle and Bioassay Columns

```{r}
nmdata_longer <- nmdata %>% 
  select(nanoparticles, Cell_Cycle_and_Proliferation, Cell_Death_Apoptosis, DNA_Damage_Repair, Cell_Adhesion_ECM, Inflammation_Immune, UPR_ER_Stress, Metal_Ion, Angiogenesis, Cytoskeleton_Organization)
```


Pivot Data Longer to have 1 response column

```{r}
nmdata_longer <- nmdata_longer %>% 
  tidyr::pivot_longer(cols = !nanoparticles, names_to = "responses", 
               values_to = "effects")
```


Group By Responses and Effects and Count 

```{r}
nmdata_longer_responses <- nmdata_longer %>% 
  group_by(responses, effects) %>% 
  count()
```


Plot Responses and Effects Count


```{r}
ggplot(data = nmdata_longer_responses, aes(fill = effects, y=n, x=responses)) + 
    geom_bar(position="dodge", stat="identity") + 
  coord_flip() + 
  theme_minimal() +
  scale_fill_brewer(palette = "Dark2", direction = -1)
```


Group by Nanoparticles and Effects and Count 

```{r}
nmdata_longer_nanoparticles <- nmdata_longer %>% 
  group_by(nanoparticles, effects) %>% 
  count()
```


Plot Nanoparticles and Effects Count

```{r}
ggplot(data = nmdata_longer_nanoparticles, aes(fill = effects, y=n, x=nanoparticles)) + 
    geom_bar(position="dodge", stat="identity") + 
  coord_flip() + 
  theme_minimal() +
    scale_fill_brewer(palette = "Dark2", direction = -1)
```

Group by Nanoparticles, Responses, and Effects and Count

```{r}
nmdata_longer_nanoparticles_effects <- nmdata_longer %>% 
  group_by(nanoparticles, responses, effects) %>% 
  count()
```


Plot Nanoparticles, Responses, and Effects Count

```{r}
ggplot(data = nmdata_longer_nanoparticles_effects,                         # Draw barplot with grouping & stacking
       aes(x = n,
           y = responses,
           fill = nanoparticles)) + 
  geom_bar(stat = "identity",
           position = "stack") +
  facet_grid(~ effects) + 
  #coord_flip() + 
  theme_minimal() +
  scale_fill_brewer(palette="Set3") 

```
Violin Plot

Select Columns of Interest

```{r}
nmdata_longer_dose_duration <- nmdata %>% 
  select(nanoparticles, UPR_ER_Stress, Angiogenesis, Metal_Ion, Inflammation_Immune, Cell_Adhesion_ECM, DNA_Damage_Repair, Cell_Death_Apoptosis, Cell_Cycle_and_Proliferation, Cytoskeleton_Organization, exposure_dose_mg_ml, exposure_duration_h)
```

Pivot Data Longer to have 1 response column

```{r}
nmdata_longer_dose_duration <- nmdata_longer_dose_duration %>% 
  tidyr::pivot_longer(cols = (UPR_ER_Stress: Cytoskeleton_Organization), 
                      names_to = "responses", 
               values_to = "effects")
```


Change Exposure dose to numeric class

```{r}
nmdata_longer_dose_duration$exposure_dose_mg_ml <- as.numeric(nmdata_longer_dose_duration$exposure_dose_mg_ml)
```

Plot Violin Plot


```{r}
ggplot(data = nmdata_longer_dose_duration, aes(fill=effects, y=exposure_dose_mg_ml, x=responses)) + 
    geom_violin(position="dodge", alpha=0.5, outlier.colour="transparent") +
    scale_fill_viridis(discrete=T, name="") +
    theme_ipsum()  +
    xlab("Bioassay") +
    ylab("exposure dose mg/mL") +
    ggtitle("Influence of Nanoparticle Dose on Bioassay Outcome up to 30 mg/mL") +
  theme(plot.title = element_text(size = 10, hjust = 0.5)) +
    ylim(0,30) +
  coord_flip() 
```
