---
title: "random_forest_classification"
author: "Connor Flynn"
date: "10/13/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load libraries
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}

librarian::shelf(here,
readr,
janitor,
tidyr,
tidyverse,
plotly,
randomForest,
edarf,
randomForestExplainer)

```

The purpose of this R Markdown is to utilize a random forest classification algorithm to demonstrate the most important factors in predicting whether a bioassay will be triggered or not


Load in Data


```{r}
#Import data
nmdata <- read_csv(here("data/Literature_transcriptomics_review_and_data_of_Nanoparticle_Induced_Cellular_Outcomes.csv"))

```

Clean Column Names


```{r}
nmdata <- nmdata %>% 
  clean_names()
```

Replace "?" with NA Values


```{r}
nmdata[nmdata == "?"] <- NA
```

Omit rows with NA Values (Dataset goes from 244 observations to 178)

```{r}
nmdata_no_na <- na.omit(nmdata)
```



Convert response column (column model is guessing) to class factor
```{r}
#cell cycle and profileration 
nmdata_no_na$cell_cycle_and_proliferation_responses <- as.factor(nmdata_no_na$cell_cycle_and_proliferation_responses)

#cell death and aoptosis
nmdata_no_na$cell_death_and_apoptosis_responses <- as.factor(nmdata_no_na$cell_death_and_apoptosis_responses)

#dna damage and repair
nmdata_no_na$dna_damage_and_repair_responses <- as.factor(nmdata_no_na$dna_damage_and_repair_responses)

#cell adhesion and extracelluar matrix response
nmdata_no_na$cell_adhesion_and_extracellular_matrix_responses <- as.factor(nmdata_no_na$cell_adhesion_and_extracellular_matrix_responses)

#inflammation and immune responses
nmdata_no_na$inflammation_and_immune_responses<- as.factor(nmdata_no_na$inflammation_and_immune_responses)

#protein responses and endoplastimic reticulum er stress
nmdata_no_na$unfolded_protein_responses_upr_and_endoplasmatic_reticulum_er_stress <- as.factor(nmdata_no_na$unfolded_protein_responses_upr_and_endoplasmatic_reticulum_er_stress)

#metal ion responses
nmdata_no_na$metal_ion_responses <- as.factor(nmdata_no_na$metal_ion_responses)

#angio genesis responses
nmdata_no_na$angiogenenesis_responses <- as.factor(nmdata_no_na$angiogenenesis_responses)

#cytoskeleton organization responses
nmdata_no_na$cytoskeleton_organization_responses<- as.factor(nmdata_no_na$cytoskeleton_organization_responses)


```


Create random forest model to predict triggered vs no effect in cell_cycle_and_proliferation_responses
```{r}
rf_cell_cycle_and_proliferation_responses <- randomForest(cell_cycle_and_proliferation_responses ~ ., 
                       data = nmdata_no_na, 
                       importance = TRUE,
                       proximity = TRUE)

rf_cell_cycle_and_proliferation_responses 

```

```{r}
#Create random forest model to predict triggered vs no effect in cell_death_and_apoptosis_responses
rf_cell_death_and_apoptosis_responses <- randomForest(cell_death_and_apoptosis_responses ~ ., 
                       data = nmdata_no_na, 
                       importance = TRUE,
                       proximity = TRUE)

rf_cell_death_and_apoptosis_responses

```

```{r}
#Create random forest model to predict triggered vs no effect in dna_damage_and_repair_responses

rf_dna_damage_and_repair_responses <- randomForest(dna_damage_and_repair_responses ~ ., 
                       data = nmdata_no_na, 
                       importance = TRUE,
                       proximity = TRUE)


rf_dna_damage_and_repair_responses
```

```{r}
#Create random forest model to predict triggered vs no effect in cell_adhesion_and_extracellular_matrix_responses

rf_cell_adhesion_and_extracellular_matrix_responses <- randomForest(cell_adhesion_and_extracellular_matrix_responses ~ ., 
                       data = nmdata_no_na, 
                       importance = TRUE,
                       proximity = TRUE)

rf_cell_adhesion_and_extracellular_matrix_responses


```

```{r}

#Create random forest model to predict triggered vs no effect in inflammation_and_immune_responses

rf_inflammation_and_immune_responses <- randomForest(inflammation_and_immune_responses ~ ., 
                       data = nmdata_no_na, 
                       importance = TRUE,
                       proximity = TRUE)

rf_inflammation_and_immune_responses
```


```{r}

#Create random forest model to predict triggered vs no effect in unfolded_protein_responses_upr_and_endoplasmatic_reticulum_er_stress

rf_unfolded_protein_responses_upr_and_endoplasmatic_reticulum_er_stress <- randomForest(unfolded_protein_responses_upr_and_endoplasmatic_reticulum_er_stress ~ ., 
                       data = nmdata_no_na, 
                       importance = TRUE,
                       proximity = TRUE)

rf_unfolded_protein_responses_upr_and_endoplasmatic_reticulum_er_stress

```


```{r}
#Create random forest model to predict triggered vs no effect in metal_ion_responses

rf_metal_ion_responses <- randomForest(metal_ion_responses ~ ., 
                       data = nmdata_no_na, 
                       importance = TRUE,
                       proximity = TRUE)


rf_metal_ion_responses

```


```{r}
#Create random forest model to predict triggered vs no effect in angiogenenesis_responses

rf_angiogenenesis_responses <- randomForest(angiogenenesis_responses ~ ., 
                       data = nmdata_no_na, 
                       importance = TRUE,
                       proximity = TRUE)

rf_angiogenenesis_responses

```


```{r}
#Create random forest model to predict triggered vs no effect in angiogenenesis_responses

rf_cytoskeleton_organization_responses <- randomForest(cytoskeleton_organization_responses ~ ., 
                       data = nmdata_no_na, 
                       importance = TRUE,
                       proximity = TRUE)


rf_cytoskeleton_organization_responses

```





Model Accuracy 
```{r}
rf_cell_cycle_and_proliferation_responses
```


Create Variable Importance Visualization

```{r}

# cell death and apoptosis variables of importance
ImpData_rf_cell_death_and_apoptosis_responses <- as.data.frame(importance(rf_cell_death_and_apoptosis_responses))
ImpData_rf_cell_death_and_apoptosis_responses$Var.Names <- row.names(ImpData_rf_cell_death_and_apoptosis_responses)

ImpData_rf_cell_death_and_apoptosis_responses <- ImpData_rf_cell_death_and_apoptosis_responses %>% mutate(Var.Names = fct_reorder(Var.Names, .[['MeanDecreaseAccuracy']]))

ggplot(ImpData_rf_cell_death_and_apoptosis_responses, aes(x=Var.Names, y=`MeanDecreaseAccuracy`)) +
  geom_segment( aes(x=Var.Names, xend=Var.Names, y=0, yend=`MeanDecreaseAccuracy`), color="skyblue") +
  geom_point(aes(size = MeanDecreaseGini), color="blue", alpha=0.6) +
  theme_light() +
  coord_flip() +
  theme(
    legend.position="bottom",
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank()
  )

```
```{r}

#cell cycle and proliferation variable of importances
ImpData_rf_cell_cycle_and_proliferation_responses <- as.data.frame(importance(rf_cell_cycle_and_proliferation_responses))
ImpData_rf_cell_cycle_and_proliferation_responses$Var.Names <- row.names(ImpData_rf_cell_cycle_and_proliferation_responses)

ImpData_rf_cell_cycle_and_proliferation_responses <- ImpData_rf_cell_cycle_and_proliferation_responses %>% mutate(Var.Names = fct_reorder(Var.Names, .[['MeanDecreaseAccuracy']]))

ggplot(ImpData_rf_cell_cycle_and_proliferation_responses, aes(x=Var.Names, y=`MeanDecreaseAccuracy`)) +
  geom_segment( aes(x=Var.Names, xend=Var.Names, y=0, yend=`MeanDecreaseAccuracy`), color="skyblue") +
  geom_point(aes(size = MeanDecreaseGini), color="blue", alpha=0.6) +
  theme_light() +
  coord_flip() +
  theme(
    legend.position="bottom",
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank()
  )

```



```{r}
#Dna_damage_and_repair_responses variable of importance
ImpData_rf_dna_damage_and_repair_responses <- as.data.frame(importance(rf_dna_damage_and_repair_responses))
ImpData_rf_dna_damage_and_repair_responses$Var.Names <- row.names(ImpData_rf_dna_damage_and_repair_responses)

ImpData_rf_dna_damage_and_repair_responses <- ImpData_rf_dna_damage_and_repair_responses %>% mutate(Var.Names = fct_reorder(Var.Names, .[['MeanDecreaseAccuracy']]))

ggplot(ImpData_rf_dna_damage_and_repair_responses, aes(x=Var.Names, y=`MeanDecreaseAccuracy`)) +
  geom_segment( aes(x=Var.Names, xend=Var.Names, y=0, yend=`MeanDecreaseAccuracy`), color="skyblue") +
  geom_point(aes(size = MeanDecreaseGini), color="blue", alpha=0.6) +
  theme_light() +
  coord_flip() +
  theme(
    legend.position="bottom",
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank()
  )

```


```{r}
# cell_adhesion_and_extracellular_matrix_responses variables of importance
ImpData_rf_cell_adhesion_and_extracellular_matrix_responses <- as.data.frame(importance(rf_cell_adhesion_and_extracellular_matrix_responses))
ImpData_rf_cell_adhesion_and_extracellular_matrix_responses$Var.Names <- row.names(ImpData_rf_cell_adhesion_and_extracellular_matrix_responses)

ImpData_rf_cell_adhesion_and_extracellular_matrix_responses <- ImpData_rf_cell_adhesion_and_extracellular_matrix_responses %>% mutate(Var.Names = fct_reorder(Var.Names, .[['MeanDecreaseAccuracy']]))

ggplot(ImpData_rf_cell_adhesion_and_extracellular_matrix_responses, aes(x=Var.Names, y=`MeanDecreaseAccuracy`)) +
  geom_segment( aes(x=Var.Names, xend=Var.Names, y=0, yend=`MeanDecreaseAccuracy`), color="skyblue") +
  geom_point(aes(size = MeanDecreaseGini), color="blue", alpha=0.6) +
  theme_light() +
  coord_flip() +
  theme(
    legend.position="bottom",
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank()
  )



```



```{r}

#inflammation_and_immune_responses variable of importance
ImpData_rf_inflammation_and_immune_responses <- as.data.frame(importance(rf_inflammation_and_immune_responses))
ImpData_rf_inflammation_and_immune_responses$Var.Names <- row.names(ImpData_rf_inflammation_and_immune_responses)

ImpData_rf_inflammation_and_immune_responses <- ImpData_rf_inflammation_and_immune_responses %>% mutate(Var.Names = fct_reorder(Var.Names, .[['MeanDecreaseAccuracy']]))

ggplot(ImpData_rf_inflammation_and_immune_responses, aes(x=Var.Names, y=`MeanDecreaseAccuracy`)) +
  geom_segment( aes(x=Var.Names, xend=Var.Names, y=0, yend=`MeanDecreaseAccuracy`), color="skyblue") +
  geom_point(aes(size = MeanDecreaseGini), color="blue", alpha=0.6) +
  theme_light() +
  coord_flip() +
  theme(
    legend.position="bottom",
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank()
  )


```

```{r}

#unfolded_protein_responses_upr_and_endoplasmatic_reticulum_er_stress variable of importance
ImpData_rf_unfolded_protein_responses_upr_and_endoplasmatic_reticulum_er_stress <- as.data.frame(importance(rf_unfolded_protein_responses_upr_and_endoplasmatic_reticulum_er_stress))

ImpData_rf_unfolded_protein_responses_upr_and_endoplasmatic_reticulum_er_stress$Var.Names <- row.names(ImpData_rf_unfolded_protein_responses_upr_and_endoplasmatic_reticulum_er_stress)

ImpData_rf_unfolded_protein_responses_upr_and_endoplasmatic_reticulum_er_stress <- ImpData_rf_unfolded_protein_responses_upr_and_endoplasmatic_reticulum_er_stress %>% mutate(Var.Names = fct_reorder(Var.Names, .[['MeanDecreaseAccuracy']]))

ggplot(ImpData_rf_unfolded_protein_responses_upr_and_endoplasmatic_reticulum_er_stress, aes(x=Var.Names, y=`MeanDecreaseAccuracy`)) +
  geom_segment( aes(x=Var.Names, xend=Var.Names, y=0, yend=`MeanDecreaseAccuracy`), color="skyblue") +
  geom_point(aes(size = MeanDecreaseGini), color="blue", alpha=0.6) +
  theme_light() +
  coord_flip() +
  theme(
    legend.position="bottom",
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank()
  )


```

```{r}
#metal_ion_responses variable of importance
ImpData_rf_metal_ion_responses <- as.data.frame(importance(rf_metal_ion_responses))
ImpData_rf_metal_ion_responses$Var.Names <- row.names(ImpData_rf_metal_ion_responses)

ImpData_rf_metal_ion_responses <- ImpData_rf_metal_ion_responses %>% mutate(Var.Names = fct_reorder(Var.Names, .[['MeanDecreaseAccuracy']]))

ggplot(ImpData_rf_metal_ion_responses, aes(x=Var.Names, y=`MeanDecreaseAccuracy`)) +
  geom_segment( aes(x=Var.Names, xend=Var.Names, y=0, yend=`MeanDecreaseAccuracy`), color="skyblue") +
  geom_point(aes(size = MeanDecreaseGini), color="blue", alpha=0.6) +
  theme_light() +
  coord_flip() +
  theme(
    legend.position="bottom",
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank()
  )

```


```{r}
#cytoskeleton_organization_responses variable of importance
ImpData_rf_cytoskeleton_organization_responses <- as.data.frame(importance(rf_cytoskeleton_organization_responses))
ImpData_rf_cytoskeleton_organization_responses$Var.Names <- row.names(ImpData_rf_cytoskeleton_organization_responses)

ImpData_rf_cytoskeleton_organization_responses <- ImpData_rf_cytoskeleton_organization_responses %>% mutate(Var.Names = fct_reorder(Var.Names, .[['MeanDecreaseAccuracy']]))

ggplot(ImpData_rf_cytoskeleton_organization_responses, aes(x=Var.Names, y=`MeanDecreaseAccuracy`)) +
  geom_segment( aes(x=Var.Names, xend=Var.Names, y=0, yend=`MeanDecreaseAccuracy`), color="skyblue") +
  geom_point(aes(size = MeanDecreaseGini), color="blue", alpha=0.6) +
  theme_light() +
  coord_flip() +
  theme(
    legend.position="bottom",
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank()
  )

```


```{r}

#add assay column to be able to join data
ImpData_rf_metal_ion_responses$assay = "metal_ion_responses"
ImpData_rf_inflammation_and_immune_responses$assay = "inflammation_and_immune_responses"
ImpData_rf_dna_damage_and_repair_responses$assay = "dna_damage_and_repair_responses"
ImpData_rf_cytoskeleton_organization_responses$assay = "cytoskeleton_organization_responses"
ImpData_rf_cell_adhesion_and_extracellular_matrix_responses$assay = "cell_adhesion_and_extracellular_matrix_responses"
ImpData_rf_cell_death_and_apoptosis_responses$assay = "cell_death_and_apoptosis_responses"
ImpData_rf_cell_cycle_and_proliferation_responses$assay = "cell_cycle_and_proliferation_responses"
ImpData_rf_unfolded_protein_responses_upr_and_endoplasmatic_reticulum_er_stress$assay = "unfolded_protein_responses_upr_and_endoplasmatic_reticulum_er_stress"




```



Facet Wrap Plots
```{r}
#combine all data
df_list <- list(ImpData_rf_metal_ion_responses,
             ImpData_rf_inflammation_and_immune_responses, 
             ImpData_rf_dna_damage_and_repair_responses,
             ImpData_rf_cytoskeleton_organization_responses,
             ImpData_rf_cell_adhesion_and_extracellular_matrix_responses,
             ImpData_rf_cell_death_and_apoptosis_responses,
             ImpData_rf_cell_cycle_and_proliferation_responses,
             ImpData_rf_unfolded_protein_responses_upr_and_endoplasmatic_reticulum_er_stress)
```

```{r}
all_data <- all_data <- Reduce(function(x, y) merge(x, y, all=TRUE), df_list, accumulate=FALSE)

```


```{r}
plot <- ggplot(all_data, aes(x=Var.Names, y=`MeanDecreaseAccuracy`)) +
  geom_segment( aes(x=Var.Names, xend=Var.Names, y=0, yend=`MeanDecreaseAccuracy`), color="skyblue") +
  geom_point(aes(size = MeanDecreaseGini), color="blue", alpha=0.6) +
  theme_light() +
  coord_flip() +
  theme(
    legend.position="bottom",
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank()
  ) +
  facet_wrap(~assay)

ggplotly(plot)


# dna_damage_and_repair_responses, angiogenenesis_responses, cytoskeleton_organization_responses are not great models
```


Plot Well Performing Models

```{r}
#combine all data
df_list_good_models <- list(ImpData_rf_metal_ion_responses,
             ImpData_rf_inflammation_and_immune_responses, 
             ImpData_rf_cell_adhesion_and_extracellular_matrix_responses,
             ImpData_rf_cell_death_and_apoptosis_responses,
             ImpData_rf_cell_cycle_and_proliferation_responses,
             ImpData_rf_unfolded_protein_responses_upr_and_endoplasmatic_reticulum_er_stress)
```

```{r}
all_data_good_models <-  Reduce(function(x, y) merge(x, y, all=TRUE), df_list_good_models, accumulate=FALSE)

```

```{r}
plot2 <- ggplot(all_data_good_models, aes(x=Var.Names, y=`MeanDecreaseAccuracy`)) +
  geom_segment( aes(x=Var.Names, xend=Var.Names, y=0, yend=`MeanDecreaseAccuracy`), color="skyblue") +
  geom_point(aes(size = MeanDecreaseGini), color="blue", alpha=0.6) +
  theme_light() +
  coord_flip() +
  theme(
    legend.position="bottom",
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank()
  ) +
  facet_wrap(~assay)

ggplotly(plot2)


```













Create another variable importance visualization
```{r}
imp_df <- data.frame(importance(rf_cell_cycle_and_proliferation_responses, scale = FALSE, type = 1))
# Tidy up and sort the data frame
imp_df <- imp_df %>% 
  mutate(names = rownames(imp_df)) %>% 
  arrange(desc(MeanDecreaseAccuracy))
# Plot mean decreased accuracy
imp_df %>% 
  top_n(20, MeanDecreaseAccuracy) %>% 
  ggplot(aes(x = reorder(names, MeanDecreaseAccuracy),y = MeanDecreaseAccuracy)) +
  geom_col() +
  coord_flip() +
  labs(title = "Variable Importance, Sonar Dataset",
       subtitle = "Random Forests (N = 500)",
       x= "",
       y= "Mean Decrease in Accuracy") +
  theme(plot.caption = element_text(face = "italic"))

```



```{r}
df_rf <- nmdata_no_na %>% 
  mutate(predicted = predict(rf_cell_cycle_and_proliferation_responses))
```



```{r}
# Save top predictor names as character vector
nm <- as.character(imp_df$names)[1:10]
# Get partial depedence values for top predictors
pd_df <- partial_dependence(fit = rf_cell_cycle_and_proliferation_responses,
                         vars = nm,
                         data = df_rf)
                         # n = c(100, 200))
```


```{r}
# Plot partial dependence using edarf
plot_pd(pd_df)

```


```{r}
perfectPartialPlot <- function(df, x, y){
  
  # Need string for aes_string()
  centile <- "centile"
  
  # Save marginal probabilities as separate data frame
  vote_prop <- df %>% 
    select(y) %>% 
    mutate(row = row_number())
  
  # Gather predictor centiles into a single column and join vote_prop
  pd_tidy <- df %>% 
    select(x) %>% 
    gather(x, key = "predictor", value = "centile") %>% 
    na.omit() %>% 
    mutate(row = row_number()) %>% 
    left_join(vote_prop, by = "row")
  
  # Create the perfect partial plot
  ggplot(pd_tidy, aes_string(x = centile, y = y)) +
    geom_line(lwd = 1.25) +
    labs(title = "Partial Dependence",
         x = "",
         y = paste("Proportion of votes for", y)) +
    facet_wrap(~predictor, scale = "free") +
    scale_y_continuous(breaks = scales::pretty_breaks(n = 4)) +
    theme(plot.title = element_text(hjust = 0.5))
  
}
```



```{r}
perfectPartialPlot(df = pd_df, x = nm, y = "triggered")
```

Random Forest Explainer

Cell Cycle and proliferation

```{r}
explain_forest(rf_cell_cycle_and_proliferation_responses, interactions = TRUE, data = nmdata_no_na, path = here("random_forests_explained/cell_cycle_and_proliferation_explained.html"))
```



```{r}
plot_predict_interaction(rf_cell_cycle_and_proliferation_responses, nmdata_no_na, "exposure_dose_mg_ml", "zeta_potential_m_v")
```

```{r}
micro <- nmdata_no_na %>% 
  group_by(nanoparticles, microarray_method_transcriptomics, cell_cycle_and_proliferation_responses) %>% 
  count
```


Can't seem to find another continuous variable that is important with zeta potential regarding prediction interaction

Metal Ion Responses

```{r}
explain_forest(rf_metal_ion_responses, interactions = TRUE, data = nmdata_no_na, path = here("random_forests_explained/metal_ion_explained.html"))
```




Cell Adhesion and Extra cellular Matrix Responses

```{r}
explain_forest(rf_cell_adhesion_and_extracellular_matrix_responses, interactions = TRUE, data = nmdata_no_na, path = here("random_forests_explained/cell_adhesion_and_extracellular_matrix_explained.html"))
```



Cell Death and Apoptosis Responses

```{r}
explain_forest(rf_cell_death_and_apoptosis_responses, interactions = TRUE, data = nmdata_no_na, path = here("random_forests_explained/cell_death_and_apoptosis_explained.html"))
```


Inflammation and Immune Responses

```{r}
explain_forest(rf_inflammation_and_immune_responses, interactions = TRUE, data = nmdata_no_na, path = here("random_forests_explained/inflammation_and_immune_explained.html"))
```


Unfolded Protein and Endoplasmatic Reticulum Stress

```{r}
explain_forest(rf_unfolded_protein_responses_upr_and_endoplasmatic_reticulum_er_stress, interactions = TRUE, data = nmdata_no_na, path = here("random_forests_explained/unfolded_protein_responses_upr_and_endoplasmatic_reticulum_er_stress_explained.html"))
```


